{\LARGE Basis Choice for Scalar-on-Function Regression \\ with Applications to Near-Infrared Spectroscopy}

### 2022-02-08 Paper/main.tex
Finishing up first take on results

	\paragraph{Monomial Basis}
	Due to the in the number of basis functions increasing collinearity of these basis functions, simulations were conducted up until the sixth monomial basis, which already shows signs of numerical instability. For reasons outlined in {\color{green} Link chapter}, they seem suited for $f_1$. The monomial basis shows a better performance than B-splines for this coefficient function. A hypothesis for this could be that $f_1$ is an entire function, which can be well approximated with a power series. In case of $f_2$, this basis shows the weakest performance out of all basis, which is visible in \hyperref[basis_expansion_2_1]{figure 9} and \hyperref[basis_expansion_2_2]{10}, where it seems like $\hat{\beta}(t)$ is not changing in the amount of noise and shows exaggerated behaviour at the boundaries. This weakness is especially pronounced in the MSPE for $f_2,Y_1$. For $f_1$, the simulation selects 5(3) and for $f_2$ 5(5) monomial basis functions for the high(low) signal-to-noise ratios. 
	\vspace{-0.2cm}

	\paragraph{B-spline Basis}
	Simulations with B-spline basis functions were possible from 4 to 18, since from 18 onwards the simulations were running into collinearity problems.
	Function $f_1$ requires only 5(4) B-spline basis functions, for the high(low) signal-to-noise ratio to obtain the best fit for the B-spline basis, which performs the worst for $f_1$. An explanation might be its exaggerated behaviour at the boundaries and the exaggeration of the peculiarities of $f_1$ (\hyperref[basis_expansion_1_1]{ADD} and \hyperref[basis_expansion_1_2]{ADD}), which is especially pronounced for the higher noise responses. For $f_2$, 11(6) B-spline basis functions are chosen for the high(low) signal-to-noise ratios. For $f_2$, the B-spline basis outperforms the monomial basis but comes second to the Fourier basis. While the basis seems to recognize the peculiarities in $f_2,Y_1$, it seems to struggle with it for the noisy responses in $f_2,Y_2$ ((\hyperref[basis_expansion_2_1]{ADD} and \hyperref[basis_expansion_2_2]{ADD})).
	\vspace{-0.2cm}

@@

	\subsubsection{Functional Principal Component Regression}
		The model which is used for the FPCR is described in \hyperref[fpc_exp_transf]{Estimation using Functional Principal Components}. Additionally to the selection of the truncation parameter $L$, the choice of the number of FPC's adds to the complexity of the model since the approximated eigenvalues and -functions from the decomposition are influenced by the choice of $L$ that was used to approximate the function. But since the FPCR is ultimately estimated with a linear model with the FPC's as regressors, the relevant degrees of freedom in the estimation of the ultimate model are not affected by $K$, but only by the number of FPC's $n_{FPC} \in \{2, 3, 4 \}$. As will be described, it seems that neither a convex behavior of the MSPE, nor a clear relationship can be observed between the increasing number of basis function and the number of FPC's. This might be since the dependency (FPCA is performed on the smoothed curves) is to complex to draw conclusions from this simulation study, therefore we will limit ourself to a brief description, The results indicate that the FPCR in this simulation seems to not take into account the differing signal-to-noise ratios of the responses since the FPC's, which affect the relevant degrees of freedom, are calculated solely from the smoothed curves, not considering the noise in the responses at all.

	\paragraph{Monomial Basis}
	For $f_1$, the cross-validated MSPE is decreasing in the increasing number of FPC's and chosen basis functions for both the high, and the low signal-to-noise ratio (4,5,6 basis functions for $n_{FPC} \in \{2, 3, 4 \}$). In  $f_2$ we also observe a decreasing MSPE in the number of FPC, but no clear relationship for the choosen basis functions. The monomial basis shows the weakest performance for all three basis functions in each setting.
	\vspace{-0.2cm}

	\paragraph{B-spline Basis}
	For $f_1,Y_1$, the MSPE suggests a better model with a higher number of FPC's. While the number of basis functions stays at five for $n_{FPC} = 2,3$, it increases to 6 basis functions for $n_{FPC} = 4$. In $f_1,Y_2$, the cross-validated MSPE is only slightly affected by $n_{FPC}$, but lowest for $n_{FPC}$ = 3, indicating the possibility of over-fitting for $n_{FPC}$ = 4. Similar to the two setups with $f_1$, the cross-validated number of basis functions for $f_2$ is increasing in the number of FPC's (4,6,23 for $n_{FPC} \in \{2, 3, 4 \}$)
	\vspace{-0.2cm}

	\paragraph{Fourier Basis}
	In $f_1,Y_1$, the MSPE is decreasing in $n_{FPC}$ while the results in $f_1,Y_2$ might indicate overfitting when using four FPC's. $f_2,Y_1$ displays the greatest relative decline of MSE in the increasing number of FPC's, but in absolute values, therefore acknowledging the higher noise responses for $f_2,Y2$, the decline of MSPE is similar to the decline observed for $f_2,Y_2$. Both configurations of $f_2$ are using 5,15,7 basis functions for $n_{FPC} \in \{2, 3, 4 \}$.
	\vspace{-0.2cm}

	\subsubsection{Interpretation and Relevance for Application}
	A possible explanation applicable for the setups performing Basis Expansion Regression might be the effect of the Bias-Variance tradeoff and the following hypothesis: For $f_1$ it might be that little bias is introduced when choosing a small number of basis functions. For $f_2$, a higher number of basis functions seems appropriate, resulting from the inherent peculiarities of $f_2$. This results in higher numbers of basis functions since the amount of bias is decreasing faster in the number of basis functions than the variance is increasing compared to $f_1$. The uncovered and described convex behaviour of the MSPE also might partly attributed by the bias-variance tradeoff. This convex behavior was not observed for the FPCR where no clear relationship between the basis functions and the number of FPC could be uncovered. An hypothesis for this is that the dependency between the truncation $L$ and the FPC's, which are calculated from $L$-truncated smoothed curves, is to complex to capture here. Following this, it appears that the results for the FPCR are mostly driven by the selected FPC´s and less by the choice of basis. To examine this hypothesis, additional simulations were conducted with a large truncation parameter $L \in \{50, 70\}$ for $n_{FPC} \in \{2, \dots, 7 \}$, which seem to support this claim (LINK TO APPENDIX).

@@

	%Adress problem of monomials again --> (This could be solved using orthogonal polynomials like Legendre polynomials, see \cite{Dattoli_2001})

	%Adress numerical instability?

	%maybee both in paragraph under results instead of here?

	%\subsection{Literature}
	%\begin{itemize}
	%	\item \cite{shonkwiler_explorations_2009}
	%	\item R-packages: fda, refund, mgcv
	%\end{itemize}

@@

	\subsection{Interpretation of Results}
		\subsubsection{Basis Expansion Regression}
		\paragraph{Monomial Basis} 
		The cross-validated MSPE selects 5 basis functions for the monomial basis, as three out of the four setups in the simulation study. For B-Splines, the cross-validation selects 10 basis functions and 9 basis functions for the fourier basis. Comparing the estimates in \hyperref[basis_expansion_estimate_appl]{ADD}, apart from the boundaries, B-spline and Fourier basis show similar behavior, opposing to the monomial estimate which might be attributed by the lower basis functions. However, we must exercise caution since from $L$=6 onwards, we were not able to calculate stable results for the monomial basis. Especially at the lower boundary, the monomial basis shows a strong exaggerated bahavior. The MSPE for B-splines (0.04574) and Fourier (0.04808) are similar while it is clearly knocked for the monomial basis (0.24181) by a factor of 5. In contrast to our hypothesis from the simulation study in which we partly attributed the weaker performance of the B-splines to its attributes at the boundaries, it manages to outperform the fourier basis. It is worthy to note that the reported MSPE for the B-spline and Fourier basis are close to the errors reported in the simulation study for the setup $f_2,Y_1$, which might be caused by a similar high signal-to-noise ratio of the octane numbers, but also other factors should be considered.

		\subsubsection{Functional Principal Component Regression}
		As for the simulation study, the interpretation of the results with respect to the choosen basis is difficult. Referring to the plotted estimates $\hat{\beta}(t)$ in \hyperref[FPCR2_estimate_appl]{ADD}, \hyperref[FPCR3_estimate_appl]{ADD} and \hyperref[FPCR4_estimate_appl]{ADD}, it appears that the higher $n_{FPC}$ is, the more similar behaviour of the estimate $\hat{\beta}(t)$ can be observed. Opposing to the simulation study, the number of basis functions is steadily increasing for the monomial basis. The bevahiour for the B-spline basis is similar to the one reported in the simulation study (basis functions increasing in $n_{FPC}$). In contrast to the simulation study, no evidence of overfitting was found for any of the three bases.

>>>>

	\paragraph{Monomial Basis}
	Due to the in the number of basis functions increasing collinearity of these basis functions, simulations were conducted up until the sixth monomial basis, which already shows signs of numerical instability. For reasons outlined in {\color{green} Link chapter}, they seem suited for $f_1$ where it shows a better performance than B-splines for this coefficient function. A hypothesis for this could be that $f_1$ is an entire function, which can be well approximated with a power series. In case of $f_2$, this basis shows the weakest performance out of all basis, for which the plots in \hyperref[basis_expansion_2_1]{figure 9} and \hyperref[basis_expansion_2_2]{10} provide evidence: It seems like $\hat{\beta}(t)$ is not changing in the amount of noise and shows exaggerated behaviour at the boundaries. This weakness is especially pronounced in the MSPE for $f_2,Y_1$. For $f_1$, the simulation selects 5(3) and for $f_2$ 5(5) monomial basis functions for the high(low) signal-to-noise ratio. 
	\vspace{-0.2cm}

	\paragraph{B-spline Basis}
	Simulations with B-spline basis functions were possible from 4 to 18, since from 18 onwards the simulations were running into collinearity problems.
	Function $f_1$ requires 5(4) B-spline basis functions, for the high(low) signal-to-noise ratio to obtain the best fit for the B-spline basis, which performs the worst for $f_1$. An explanation might be its exaggerated behaviour at the boundaries and the exaggeration of the peculiarities of $f_1$ (\hyperref[basis_expansion_1_1]{ADD} and \hyperref[basis_expansion_1_2]{ADD}), which is especially pronounced for the higher noise responses. For $f_2$, 11(6) B-spline basis functions are chosen for the high(low) signal-to-noise ratio. For $f_2$, the B-spline basis outperforms the monomial basis but comes second to the Fourier basis. While the basis seems to recognize the peculiarities in $f_2,Y_1$, it seems to struggle with it for the noisy responses in $f_2,Y_2$ ((\hyperref[basis_expansion_2_1]{ADD} and \hyperref[basis_expansion_2_2]{ADD})).
	\vspace{-0.2cm}

@@

	\subsubsection{Functional Principal Component Regression}
		The model which is used for the FPCR is described in \hyperref[fpc_exp_transf]{Estimation using Functional Principal Components}. Additionally to the selection of the truncation parameter $L$, the choice of the number of FPC's adds to the complexity of the model since the approximated eigenvalues and -functions from the decomposition are influenced by the choice of $L$ that was used to expand the function. But since the FPCR is ultimately estimated with a linear model with the FPC's as regressors, the relevant degrees of freedom in the estimation of the ultimate model are not affected by $K$, but only by the number of FPC's $n_{FPC} \in \{2, 3, 4 \}$. As will be described, it seems that neither a convex behavior of the MSPE, nor any clear relationship can be observed between the number of basis function and the number of FPC's. This might be because the dependency (FPCA is performed on the smoothed curves) is to complex to draw conclusions from this simulation study, therefore we will limit ourself to a brief description. In this simulation, the cross-validated choice of basis functions indicates that the FPCR might not take the differing signal-to-noise ratios of the responses into account since the FPC's, which affect the relevant degrees of freedom, are calculated solely from the smoothed curves, not considering the noise in the responses at all.

	\paragraph{Monomial Basis}
	For $f_1$, the cross-validated MSPE is decreasing in the increasing number of FPC's and chosen basis functions for both the high, and the low signal-to-noise ratio (4,5,6 basis functions for $n_{FPC} \in \{2, 3, 4 \}$). In  $f_2$ we also observe a decreasing MSPE in the number of FPC, but no clear relationship for the chosen basis functions. The monomial basis shows the weakest performance for all three basis functions in each setting for all $n_{FPC}$.
	\vspace{-0.2cm}

	\paragraph{B-spline Basis}
	For $f_1,Y_1$, the MSPE suggests better models with a higher number of FPC's. While the number of basis functions stays at five for $n_{FPC} = 2,3$, it increases to 6 basis functions for $n_{FPC} = 4$. In $f_1,Y_2$, the cross-validated MSPE is only slightly affected by $n_{FPC}$, but lowest for $n_{FPC}$ = 3, indicating the possibility of over-fitting using $n_{FPC}$ = 4. Similar to the two setups with $f_1$, the cross-validated number of basis functions for $f_2$ is increasing in the number of FPC's (4 ,6 ,23 for $n_{FPC} \in \{2, 3, 4 \}$)
	\vspace{-0.2cm}

	\paragraph{Fourier Basis}
	In $f_1,Y_1$, the MSPE is decreasing in $n_{FPC}$ while the results in $f_1,Y_2$ might indicate overfitting when using $n_{FPC}$ = 4. $f_2,Y_1$ displays the greatest relative decline of MPSE in the increasing number of FPC's, but in absolute values, therefore acknowledging the higher noise responses for $f_2,Y2$, the decline of MSPE is similar to the decline observed for $f_2,Y_2$. Both configurations of $f_2$ are using 5, 15, 7 basis functions for $n_{FPC} \in \{2, 3, 4 \}$.
	\vspace{-0.2cm}

	\subsubsection{Interpretation and Relevance for Application}
	A possible explanation applicable for the setups performing Basis Expansion Regression might be the effect of the Bias-Variance tradeoff and the following hypothesis: For $f_1$ only little bias seems introduced when choosing a small number of basis functions. For $f_2$, a higher number of basis functions seems appropriate, resulting from the inherent peculiarities of $f_2$ that are more pronounced with lower $L$. This results in higher numbers of basis functions since the amount of bias is decreasing faster in the number of basis functions than the variance is increasing compared to $f_1$. The uncovered and described convex behaviour of the MSPE also might partly attributed by the bias-variance tradeoff.

	This convex behavior was not observed for the FPCR where no clear relationship between the basis functions and the number of FPC could be uncovered. An hypothesis for this is that the dependency between the truncation $L$ and the FPC's, which are calculated from $L$-truncated smoothed curves, is to complex to capture in our setting. To examine this further, additional simulations were conducted for B-spline basis functions with a large truncation parameter $L \in \{50, 70\}$ for $n_{FPC} \in \{2, \dots, 7 \}$ (LINK TO APPENDIX). As in the simulation study we can find potential signs of overfitting for all setups and both choices of $L$. The additional simulations revealed evidence that there exists a relationship from $n_{FPC} = 3$ onwards: For $n_{FPC} \in \{3, \dots, 7 \}$, the 10-fold cross-validated MSPE is always lower than for 70 basis functions which might be a first sign of a too high $K$ and subsequent undersmoothing. Following this, it seems that once a sufficient number of basis functions is used to expand the curves, the FPCR performs better on a lower truncated smoothed curve (here $L$ = 50). Importantly, the settings $f_1,Y_2$ and $f_2,Y_2$ perform worse for these high number of basis functions, while $f_1,Y_1$ and $f_2,Y_1$ show a stronger performance than in the main simulation study, although the signal-to-noise ration of the responses is not affecting the construction of the FPC's which might indicate benefiting attributes of the lower basis functions used in the main simulation.

@@

@@

	\subsection{Interpretation of Results}
		\subsubsection{Basis Expansion Regression}
		The cross-validated MSPE selects 5 basis functions for the monomial basis. For B-Splines, the cross-validation selects 10 basis functions and 9 basis functions for the fourier basis. Comparing the estimates in \hyperref[basis_expansion_estimate_appl]{ADD}, apart from the boundaries, B-spline and Fourier basis show similar behavior, opposing to the monomial estimate which might be attributed by the lower amount of used basis functions. Especially at the lower boundary, the monomial basis shows a strong exaggerated bahavior. However, we must exercise caution since from $L$=6 onwards, we were not able to calculate stable results for the monomial basis. The MSPE for B-splines (0.04574) and Fourier (0.04808) are similar while it is clearly knocked of for the monomial basis (0.24181). In contrast to the simulation study, the B-splines basis manages to outperform the fourier basis. Driving factors for this better performance could be that first, we do not assume a periodic true coefficient function with the same start- and end value and second, the Fourier basis tries to force the same start- and endvalue on $NIR$ (confirmed in unreported visual inspection). It is worthy to note that the reported MSPE for the B-spline and Fourier basis are close to the errors reported in the simulation study for the setup $f_2,Y_1$, which might be caused by the mentioned similarities between $NIR$ and $f_2$, but also by a similar high signal-to-noise ratio of the octane numbers.

		\subsubsection{Functional Principal Component Regression}
		The best performance for all numbers of FPC's was achieved with the Fourier Basis, as for the FPCR in the simulation study. In contrast to the simulation study, no evidence of overfitting was found for the best basis choices in any of the three bases. The MSPE appears to be strictly decreasing in the number of FPC´s for its best basis choice. As for the simulation study, the interpretation of the results with respect to the chosen basis is difficult: Referring to the plotted estimates $\hat{\beta}(t)$ in \hyperref[FPCR2_estimate_appl]{ADD}, \hyperref[FPCR3_estimate_appl]{ADD} and \hyperref[FPCR4_estimate_appl]{ADD}, it appears that the higher $n_{FPC}$ is, the more similar behaviour of the estimate $\hat{\beta}(t)$ can be observed. Opposing to the simulation study, the number of basis functions is steadily increasing for the monomial basis. The behavior for the B-spline basis is similar to the one reported in the simulation study (basis functions increasing in $n_{FPC}$).
